Lie Algebras and Groups\vspace{-1ex}

### 2019-02-13 Lie Algebra Notes/Lie Algebras.tex
I should push more frequently

@@

%proofs
\theoremstyle{nonumberbreak}
\theoremindent0.5cm
\theoremheaderfont{\sc}
\theoremseparator{}
\theoremsymbol{\ensuremath\spadesuit}
\newtheorem{prf}{Proof}

@@

%Image
\DeclareMathOperator{\im}{Im}

@@

%Lie algebra stuff
\DeclareMathOperator{\gl}{\mathfrak{gl}}
\let\sl\relax
\DeclareMathOperator{\sl}{\mathfrak{sl}}
\DeclareMathOperator{\so}{\mathfrak{so}}
\let\sp\relax
\DeclareMathOperator{\sp}{\mathfrak{sp}}
\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\ad}{ad}
\DeclareMathOperator{\Der}{Der}
\DeclareMathOperator{\Int}{Int}
\DeclareMathOperator{\Rad}{Rad}
\let\t\relax
\DeclareMathOperator{\t}{\mathfrak{t}}
\DeclareMathOperator{\n}{\mathfrak{n}}
\DeclareMathOperator{\diag}{diag}

@@

	First, notice we have
	\[[e_{i,i+1}e_{i+1,i}]=e_{ii}-e_{i+1,i+1},\]
	so it suffices to show we also have the $e_{ij}$ for $i\ne j$.

\end{sol}

@@

\begin{hwprob}{3.8}
	Let $L$ be nilpotent. Prove that $L$ has an ideal of codimension 1.
\end{hwprob}
\begin{sol}
	Assume that $L$ is nilpotent. Then defining $L^0=L$, and by induction $L^i=[LL^{i-1}]$, this 
	means that $L^n=0$.
\end{sol}

\begin{hwprob}{3.9}
	Prove that every nilpotent Lie algebra L has an outer derivation (see 1.3) as follows: Write $L=K+Fx$
	for some ideal $K$ of codimension 1. Then $C_L(K)\ne 0$ (why?). Choose $n$ so that $C_L(K)\subseteq L^n$,
	$C_L(K)\not\subseteq L^{n+1}$ and let $z\in C_L(K)-L^{n+1}$. Then the linear map $\delta$ sending $K$ to zero
	and $x$ to $z$ is an outer derivation.
\end{hwprob}
\begin{sol}

\end{sol}

\begin{hwprob}{4.3}
	This exercise illustrates the failure of Lie's theorem when $F$ is allowed to have positive characteristic.
	Consider the $p\times p$ matrices
	\[x=\begin{pmatrix} 
		0 & 1 & 0 & \cdots & 0\\
		0 & 0 & 1 & \cdots & 0\\
		\vdots & \vdots & \vdots & \ddots & \vdots\\
		0 & 0 & 0 & \cdots & 1\\
		1 & 0 & 0 & \cdots & 0
	\end{pmatrix}\quad y=\diag(0,1,\dots, p-1)\]
	Check that $[xy]=x$, hence that $x$ and $y$ span a two dimensional solvable subalgebra $L$ of $\gl(p,F)$. Verify that $x$ and $y$ have no common eigenvector.
\end{hwprob}
\begin{sol}

\end{sol}
\newpage
%%%% End HW 1 %%%%

@@

\end{document}

>>>>

%Shade definitions
\theoremindent0cm
\theoremheaderfont{\normalfont\bfseries} 
\def\theoremframecommand{\colorbox[rgb]{.9,.9,.9}}
\newshadedtheorem{ex}[thm]{Example}

@@

%proofs
\theoremstyle{nonumberbreak}
\theoremindent0.5cm
\theoremheaderfont{\sc}
\theoremseparator{}
\theoremsymbol{\ensuremath\spadesuit}
\newtheorem{prf}{Proof}
\newtheorem{conj}{Conjecture}

@@

%Image
\DeclareMathOperator{\im}{im}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\End}{End}

@@

%Lie algebra stuff
\DeclareMathOperator{\gl}{\mathfrak{gl}}
\let\sl\relax
\DeclareMathOperator{\sl}{\mathfrak{sl}}
\DeclareMathOperator{\so}{\mathfrak{so}}
\let\sp\relax
\DeclareMathOperator{\sp}{\mathfrak{sp}}
\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\ad}{ad}
\DeclareMathOperator{\Der}{Der}
\DeclareMathOperator{\Int}{Int}
\DeclareMathOperator{\Rad}{Rad}
\let\t\relax
\DeclareMathOperator{\t}{\mathfrak{t}}
\DeclareMathOperator{\n}{\mathfrak{n}}
\DeclareMathOperator{\h}{\mathfrak{h}}
\DeclareMathOperator{\g}{\mathfrak{g}}
\DeclareMathOperator{\diag}{diag}

@@

	First, notice we have
	\[[e_{i,i+1}e_{i+1,i}]=e_{ii}-e_{i+1,i+1},\]
	so it suffices to show we also have the $e_{ij}$ for $i\ne j$.
	This is clear whenever $k\ge 2$ (so that we have three columns in our matrices). Let $i\ne j$ be arbitrary
	and $l$ be a (the) third index. Then 
	\[e_{il}e_{lj}]=e_{ij}.\]

	If $k=1,$ then 
	\[e_{21},e_{11}-e_{22}]=e_{21}+e_{21}=2e_{21}\]
	and so $e_{21}$ is in the space generated by $[\calB \calB]$ and an analagous proof
	gives us the rest of the basis.

	\subsubsection*{Type $B_k$:}
	This one is getting a bit long and is low on my priority list so might not be finished.
\end{sol}

@@

\begin{hwprob}{3.8}
	Let $L$ be nilpotent. Prove that $L$ has an ideal of codimension 1.
\end{hwprob}
\begin{sol}
	Assume that $L$ is nilpotent. That is, defining $L^0=L$, and by induction $L^i=[LL^{i-1}]$, this 
	means that $L^n=0$.

	We proceed by induction on $\dim L$: for the base case, if $L=0$ the result doesn't really mean anything,
	and when $\dim L=1$, the trivial subspace has codimension 1.

	Assume now that $\dim L=i$ and that every Lie algebra of dimension less than $i$ has a codimension 1 ideal. 
	Then we have two subcases to consider. If $L$ is commutative then
	every vector subspace is an ideal, so take a codimension one subspace and we are finished.

	Finally assume $[LL]\ne 0$ and consider the quotient $Q=L/[LL]$. By 
	our assumptions (and since $L$ is nilpotent $L\ne[LL]$) we get that
	\[0<\dim Q<\dim L\]
	and that, in fact, by rank-nullity
	\[\dim Q = \dim L - \dim[LL].\]
	But then by the induction hypothesis there is an ideal $J\lhd Q$ with dimension $\dim Q - 1$
	and pulling this back through the quotient yields the ideal $\bar J\lhd L$ that has dimension $\dim Q-1+\dim[LL]=\dim L-1$. 
	This is our codimension one ideal and we complete the proof.
\end{sol}

\begin{hwprob}{3.9}
	Prove that every nilpotent Lie algebra L has an outer derivation (see 1.3) as follows: Write $L=K+Fx$
	for some ideal $K$ of codimension 1. Then $C_L(K)\ne 0$ (why?). Choose $n$ so that $C_L(K)\subseteq L^n$,
	$C_L(K)\not\subseteq L^{n+1}$ and let $z\in C_L(K)-L^{n+1}$. Then the linear map $\delta$ sending $K$ to zero
	and $x$ to $z$ is an outer derivation.
\end{hwprob}
\begin{sol}
	Using the result above, we can write every nilpotent $L$ as $K+Fx$ for some $K\lhd L$.
	Now since $K\le L$, and since $L$ is simple, for some $n$: $L^n=[L^{n-1}L^{n-1}]=0$,
	or equivalently every $n$-fold bracket in $L$ is zero.

	Assuming that $n$ is minimal, this means that one can construct a $(n-1)$-fold bracket
	\[x=[x_1[x_2[x_3\cdots[x_{n-2}x_{n-1}]\cdots]]]\]
	which is nonzero. But then $[xL]=0$ so in particular $[xK]=0$ whence $x\in C_L(K).$

	Therefore we can choose an $m$, as suggested, such that $C_L(K)\in L^m\setminus L^{m-1}$
	and let $\delta$ be the linear map sending $K$ to zero and $x$ to some $z\in C_L(K)\setminus L^{m+1}$.
	We will now establish this is an outer derivation.

	Let $a,b\in L$ be arbitrary with $a=k_a+l_ax$ and $b=k_b+l_bx$ and compute
	\begin{align*}
		\delta(ab) &= \delta(k_a+l_ax)(k_b+l_bx)\\
		&= \delta(k_ak_b+k_al_bx+l_axk_b+l_axl_bx)\\
		&=0+k_a\delta(l_bx)+\delta(l_ax)k_b+\delta(l_ax)l_bx+l_ax\delta(l_bx)\\
		&=(k_a+l_ax)\delta(l_bx)+\delta(l_ax)(k_b+l_bx)\\
		&=a\delta(k_b+l_bx)+\delta(k_a+l_ax)b\\
		&=a\delta(b)+ \delta(a)b
	\end{align*}
	so $\delta$ is a derivation. It remains to show it is not inner.

	Assume by contradiction that it were. Then $\delta =\ad w$ for some $w\in L$. But then notice
	that $[wK]=0$ so $w\in C_L(K)\subseteq L^m$. But then bracketing with one more thing puts us in $L^{m+1}$, so
	$[w,x]\in L^{m+1}$. But then
	\[[w,x]=\ad w(x)=\delta(x)=z\in L^{m+1}\]
	a contradiction. Thus $\delta$ is an outer derivation.
\end{sol}

\begin{hwprob}{4.3}
	This exercise illustrates the failure of Lie's theorem when $F$ is allowed to have positive characteristic.
	Consider the $p\times p$ matrices
	\[x=\begin{pmatrix} 
		0 & 1 & 0 & \cdots & 0\\
		0 & 0 & 1 & \cdots & 0\\
		\vdots & \vdots & \vdots & \ddots & \vdots\\
		0 & 0 & 0 & \cdots & 1\\
		1 & 0 & 0 & \cdots & 0
	\end{pmatrix}\quad y=\diag(0,1,\dots, p-1)\]
	Check that $[xy]=x$, hence that $x$ and $y$ span a two dimensional solvable subalgebra $L$ of $\gl(p,F)$. Verify that $x$ and $y$ have no common eigenvector.
\end{hwprob}
\begin{sol}
	Compute
	\[[xy]=\begin{pmatrix} 
		0 & 1 & 0 & \cdots & 0\\
		0 & 0 & 2 & \cdots & 0\\
		\vdots & \vdots & \vdots & \ddots & \vdots\\
		0 & 0 & 0 & \cdots & p-1\\
		0 & 0 & 0 & \cdots & 0
	\end{pmatrix}-\begin{pmatrix} 
		0 & 0 & 0 & \cdots & 0\\
		0 & 0 & 1 & \cdots & 0\\
		\vdots & \vdots & \vdots & \ddots & \vdots\\
		0 & 0 & 0 & \cdots & p-2\\
		p-1 & 0 & 0 & \cdots & 0
	\end{pmatrix}=\begin{pmatrix} 
		0 & 1 & 0 & \cdots & 0\\
		0 & 0 & 1 & \cdots & 0\\
		\vdots & \vdots & \vdots & \ddots & \vdots\\
		0 & 0 & 0 & \cdots & 1\\
		1 & 0 & 0 & \cdots & 0
	\end{pmatrix}=x\]
\end{sol}
Thus if $L$ is the vector space spanned by $x$ an $y$, since it is closed under bracket, 
it is a Lie subalgebra of $\gl(p,F)$. Now because of this computation it is easy to see that
$[LL]=Fx$, so $L^2=0$, whence $L$ is solvable.

Notice that multiplication by $x$ on the right acts by permuting the columns of a matrix so $x^p=\id$,
whence the eigenvalues of $x$ are roots of $x^p-1=(x-1)^p$ since $\ch F=p$. Let $v$ me any element of $L$ 
fixed by the action of $x$ via $\ad x$ on $L$. Then
\[bx=b[x,y]=[x,ax+by]=[x,v]=xv-vx=v\]
but then since the only eigenvalue can be 1, $v=x$. But $[y,x]=-x$, which cannot be a scalar multiple of $y$,
so $x$ and $y$ can have no common eigenvector.
\newpage
%%%% End HW 1 %%%%

@@

\section{January 18, 2019}
\begin{prob}
	The ``board problem'' for this week: Show that a semisimple Lie algebra over $\C$ is generated as a Lie
	algebra by two elements.
\end{prob}

\subsection{Returning to Jordan Decomposition}
Given an $n\times n$ over an algebraically closed field $k$ of any characteristic, let $T$ be the corresponding linear
transformation on $k^n$. We know from the proof of the Jordan canonical form that 
\[V=k^n=\oplus_{a\in k}\ker(T-aI)^n\]
the sum of the \textbf{generalized eigenspaces} of $T$ on $V$. We then use the CRT:
there is a polynomial $p(T)$ such that $p(T)=a\pmod (T-a)^n$ for all eigenvalues $a$ and
$p(T)=0\pmod{T}$.

This is clear since the eigenvalues are distinct, so that all moduli are pairwise relatively prime.
SO then $p(T)$ acts on $\ker(T-A)^n$ so it is \textbf{semisimple} (diagonal) and $T-p(T)$, acting by $T-a$ on
$\ker(T-a)^n$, acts nilpotently.

Hence $p(x)$ and $x-p(x)$ are polnomials in $x$ without constant term such that as matrices they commute
and $x$ is semisimple (diagonal) and $x-p(x)$ is nilpotent. Moreover, tere is only one way to write $x$ as $s+n$ where $s$ is semisimple, $n$ is nilpotent, and $[sn]=0$.
This is since given $x=s+n$ where $s,n$ both commute with $x$, so with $p(x)$ and $x-p(x)$, we get
\[p(x)-s+(p(x)-x)+n\]
(check this) with the left side the sum of \textbf{commuting} semisimple matrices is semisimple and
the right side is the sum of \textbf{commuting} nilpotent matrices whence nilpotent, but the only matrix
which is both semisimple and nilpotent is the zero matrix.
\begin{rmk}
	Notice that commutativity above is vital to the statement.
\end{rmk}

Why is this important? We want to study the representations of Lie algebras, even the linear ones, and we would like to know that the semisimple and nilpotent parts $x_s, x_n$ of
any $x$ lying in such an algebra \textbf{continue} to act semisimply and nilotently in the representation 
(and that the parts $x_s$ and $x_n$ lie in L if $x$ does, which is true when $L$ is semisimple).

We have already seen that if $d$ is diagonal, then 
<FILL THIS IN>

Now we want to head towards a criterion in terms of traces alone for a matrix to be nilpotent.
Eventually we want to determine when a Lie algebra is solvable.

\begin{lem}
	Suppose that $k$ is algebraically closed of characteristic zero and that $B\subseteq A$ are subspaces of $\gl(n,k)$ and that $x(\in M?)$ is a matrix such that 
	$\tr(xy)=0$ for all $y\in M=\{z\in\gl(n,k)|[zA]\subseteq B\}$. Then $x$ is nilpotent.
\end{lem}
\begin{prf}
	Write $x=s+n$, the Jordan decomposition. Since $x\in M, s,n\in M$ since $s$ and $n$ are polynomials in 
	$x$ without constant term. We must show that $s=0$ -- that all eigenvalues of $s$ are zero.

	Let $a_1,\dots,a_n$ be the eigenvalues of $s$. These all lie in $k$, so they span a finite-dimensonal $\Q$-subspace $E$
	of $k$. To see this, we will show that any $\Q$-linear function $f:E\to\Q$ must be zero. Given $f$, 
	choose a polynomial $r(s)$ such that $r(a_i-a_j)=f(a_i)-f(a_j)$ for all $i$ and $j$ (using the Lagrange interpolation theorem).

	This is consistent since $a_i-a)j=a_k-a_l$ (using the linearity of $f$) implies the image of these under $r$ are equal.
	Then looking at $r(\ad s)$, we find that its eigenvalues are $r(a_i-a_j)$ as well as $f(a_i)-f(a_j)$,
	whence $\sum f(a_i)a_i=0=\tr(sy)$ for asuitable $y$.

	Applying $f$, we get $\sum f(a_i)^2=0$ whence all $f(a_i)=0$ (since $f(a_i)\in\Q$) and $f=0$, as desired.
\end{prf}

\subsection{Cartan's Criterion}
How can we turn this into something useful for us in the Lie algebra world?
\begin{thm}[Cartan's Criterion]
	Let $L$ be a linear Lie algebra over (alg. closed and char zero) $k$. Suppose that $\tr(xy)=0$
	for all $x\in L$ and $y\in[LL]$. Then $L$ is solvable.

	As a consequence, suppose that $L$ is a Lie algebra such that 
	$\tr((\ad x)(\ad y))=-$ for all $x$ and $y$ above. Then $L$ is solvable.
\end{thm}
\begin{rmk}
	Notice that the consequence really does follow from the theorem: if $\tr((\ad x)(\ad y))=0$
	then $\ad L\cong L/Z$ is solvable where $Z$ is the center of $L$ whence $L$ is also solvable
	since $Z$ is abelian.

	To prove that $L$ is solvable, under our hypothesis, it is enough to show that $[LL]$ is nilpotent, 
	but (due to Engel's theorem) it suffices to show that $[LL]$ consists of nilpotent matrices.
\end{rmk}
To prove this criterion, we need a fact about traces (the \textbf{associativity of traces}):
\[\tr([xy]z)=\tr(x[yz]).\]
\begin{prf}
	We know that $\tr(xy)=0$ for all $x\in L$, $y\in [LL]$ and any such $x$ maps $L$ to $[LL]$ under $\ad$.
	We would like to know that for any $x\in M$, that $\tr(xz)=0$ for all  $x\in[LL]$. We know this is true if $z\in L$
	so by the associativity of trace, this follows.

	Returning to the Jordan decomposition, let $A$ be any finite-dimensional algebra over $k$ (not necessarily associative)
	and let $d$ be a derivation of $A$. THen the semisimple parts $d_s$ and $d_n$ of $d$ are also derivations.

	To see this, we write $A$ as the direct sum of the generalized eigenspaces $A_a$ of itself under the action of $d$. Since $d$ is a derivation,
	it ends up becoming clear that $A_aA_b\subseteq A_{a+b}$, so that the transformation from $A$ to itself action by $a$ on $A_a$ is a derivation.
	But this is $d_s$. Then $d_n=d-d_s$ is also a derivation.
\end{prf}

\section{January 23, 2019}
Some clarifications from last time:
\begin{lem}
	Let $A\subseteq B$ be subspaces of $\gl(n,l)$ for some algebraically closed field of char zero.
	Let $M=\{x\in\gl(n,k)|[xB]\subseteq A\}$. Suppose that $x\in M$ such that $\tr(xy)=0$ for all $y\in M$.
	Then $x$ is nilpotent.
\end{lem}
\begin{rmk}
	This used the fact that if $x=x_s+x_n$ then $\ad x=\ad x_s+\ad x_n$ is the Jordan decomposition for $\ad x$.

	Furthermore, when applying this lemma to Cartan's criterion for sovability of Lie subalgebras of $\gl(n,k)$, 
	we claimed that, if $\tr(xy)=0$ for all $x\in L$ and $y\in [LL]$, then $L$ is solvable. This follows by our lemma
	since if $x\in M=\{z\in\gl(n,k)|[zL]\subseteq[LL]\}$ and if $y,z\in L$ then $\tr(x[yz])=\tr([xy]z)=0$ by hypothesis.

	So by the lemma $[LL]$ consists of nilpotents, so is nilpotent and thus solvable.
\end{rmk}

\begin{defn}
	If $L$ is any Lie algebra over $k$, then $\kappa(x,y)=\tr((\ad x)(\ad y))$, is a bilinear 
	form on $L$ and is called the \textbf{Killing form.}

	It is symmetric and associative ($\kappa([xy],z)=\kappa(x,[yz])$).
\end{defn}
\begin{rmk}
	Now in the context of \textit{general (as opposed to linear) Lie algebras} we can restate
	Cartan to say:
\end{rmk}
\begin{thm}[Cartan Redux]
	If $\kappa(x,y)$ for all $x\in [LL]$ and $y\in L$, then $L$ is solvable.
\end{thm}

\subsection{Semisimple Lie Algebras}
What we are interested in, actually, is the polar opposite of solvable algebras: namely,
\textbf{semisimple} Lie algebras, having no nonzero solvable ideals.
\begin{thm}
	$L$ is semisimple if and only if the Killing form $\kappa$ is nondegenerate.
\end{thm}
\begin{rmk}
	This sort of naturally follows from our discussion above -- we want the opposite property for our algebra
	and we require that the form be as far from zero as possible.
\end{rmk}
\begin{rmk}
	If $I\lhd L$, we note that, if $\kappa_I$ is the Killing form for $I$ and $\kappa$ the Killing form for
	$L$, then $\kappa_I=\kappa|_{I\times I}$. This is because if $x,y\in I$ then $\ad x$ and $\ad y$ both map $L$ into $I$.
	Thus if $M$ is any $n\times n$ matrix over $k$ mappting $V=k^n$ to a proper subspace $W$, then $\tr M=\tr M|_W$, since 
	extending a basis from $W$ to $M$ is such that the added basis vectors make no contribution to the trace of $M$.
\end{rmk}

\begin{prf}
	If $\kappa$ is nondegenerate and $S$ is a solvable ideal of $L$, then $\kappa|X=\kappa S$ has
	\textbf{radical} $\{z\in S:\kappa(z,y)=0,\forall y\in S\}$ which is at least $[SS]$, and then $K$ would have a 
	radical as well.

	More precisely, if $S$ is a solvable ideal of $L$ then there is a nonzero abelian ideal $A$ of $L$
	(the last nonzero term in the derived series for $L$) and then if $x\in A$ and $y\in L$, then $\ad x\ad y$ maps $L$
	into $A$ and $(\ad x\ad y)^2=0$, so $\kappa$ has radical at least $A$.

	Conversely, if $L$ is semisimple, then the radical of $\kappa$ as defined aobve would be a solvable ideal of $L$,
	by associativity and Cartan's criterion, so $\operatorname{Rad}(\kappa)=0$.
\end{prf}

The nondegeneracy of the Killing form is \textbf{the key} result that makes the classification
of semisimple Lie algebras tractable. For instance, in the case of finite simple groups the classification required 
10,000 pages of journal papers and it boils down to not having any kind of simple criterion like this.

\subsection{Applications}
As a first step, we show 
\begin{thm}
	A Lie algebra $L$ is semisimple if and only if it is a finite direct sum of simple ideals $I$.
\end{thm}
\begin{prf}
	This is clear since if $I$ is an ideal then $I^\perp=\{y\in K:\kappa(y,x)=0\forall x\in I\}$
	is again an ideal of $L$ and $I\cap I^\perp=0$ by Cartan. Then we can keep doing this, pulling off minimal ideals.
\end{prf}

As a second application, we will show that ever derivation $d$ of a semisimple Lie algebra $L$ is inner.
\begin{prf}
	$L$ is semisimple, so $L=[LL]$ and $Z(L)=0$, so $\ad L\cong L$. Now if $D=\operatorname{Der} L$, then
	$L\lhd D$ (as in homework). If $L\ne D$, then take the orthogonal of $L$ under the killing form in $D$.

	Thus $L\cap L^\perp=0$ and $\dim L^\perp\ge\dim D-\dim L$, but $[LL^\perp]=0$, which says exactly that if $m\in L^\perp$, then $m(x)=0$
	for all $x\in L$, so $L^\perp=0$ and $L=D$.
\end{prf}

Next we can extend the Jordan decomposition to semisimple Lie algebras. We have already seen that if $L$ is semisimple then $\Der L$ is closed
under Jordan decomposition but $\Der L=L$, so any $x\in L$ can be written as $x_s+x_n$ \textbf{FOR $x_s,x_n\in L$ (important)} where $\ad x_s$ and $\ad x_n$
act semisimply and nilpotently on $L$ and that $[x_sx_n]=0.$

\begin{rmk}
But what if $L$ is a semisimple linear Lie algebra over $k$? Then if $x\in L$, we have  as above $x_s,x_n\in L$
but also $x=s+n$ for some $s,n\in\gl(n,k)$ where $s$ is semisimple and $n$ is nilpotent and $[sn]=0$. Then we can ask:
do $s$ and $n$ lie in $L$? The answer is yes, but requires more machinery Once we know this, there are immediate applications to 
representation theory of $L$ (maps $\varphi:L\to\gl(n,k)$):

Given any such representation $\varphi$ and $x\in L$ with $x=x_s+x_n$, we get that $\varphi(x_s)$ and $\varphi(x_n)$
\textbf{continue to act semisimply and nilpotently} on $k^n$. This is what allows us to get a real handle on the representation theory.
\end{rmk}

\begin{thm}
	Given any \textbf{finite dimensional} module $M$ over a finite dimensional $M$ over a semisimple Lie
	algebra $L$ is completely reducible.
\end{thm}
\begin{rmk}
	We will prove this next time, using also certain constructions on modules as well as the matrix
	representations of elements.

	This result ends up being extremely useful not only in the (more obvious) representation theory of $L$, 
	but also in the structure theory of $L$ by considering the adjoint action of $L$ on itself.
\end{rmk}

\section{January 25, 2019}
\subsection{Weyl's Theorem}
TOday we'll prove Weyl's theorem:
\begin{thm}[Weyl]
	Any finite dimensional module over a semisimple Lie algebra is complete reducible ( a finite direct sum of simple modules).
\end{thm}
To prove this, we need some preparation: namely a couple of methods of producing new moduels over any Lie algebra $L$ from old ones.

First of all, if $M,N$ are $L$-modules, then $M\otimes_k N$ is an $L$ module of dimension the product
of the dimensions of $M$ and $N$. Here we define the action
\[x\cdot(m\otimes n)=x\cdot m\otimes n+m\otimes x\cdot n\]
where we can quickly check that 
\[[xy]\cdot(m\otimes n)=x(y(m\otimes n))-y(x(m\otimes n)).\]

Similarly, $\Hom_k(M,N)$ is an $L$ module via
\[(xf)(v)=xf(v)-f(xv)\]
and again we can check this defines a module structure, noting that $xf=0$ for all $x\in L$ if and only if $f$ is an $L$ module homomorphism.

\begin{lem}[Shur's Lemma]
	If $M$ is an irreducible $L$ module, then the only $L$ module maps in $\operatorname{End}(M)$ are scalars.
\end{lem}

Last we need
\begin{defn}
	We have the \textbf{trace form} on $L$:
	\[(x,y)=(\tr\varphi(x))(\tr\varphi(y))\]
	and by associativity and Cartan's criterion, we know (assuming that $L$ acts faithfully) that $(-,-)$ is nondegenerate.
\end{defn}
Recall that given any basis $x_i$ of $L$, we can define the dual basis $y_i$ such that $(x_i,y_j)=\delta_{ij}$.
\begin{defn}
	The \textbf{Casimir element} attached to a homomorphism $\varphi:L\to\gl(n,k)$, for some $n$, to be the matrix
	\[c_\varphi=\sum_1^m\varphi(x_i)\varphi(y_i).\]
\end{defn}
\begin{rmk}
	This matrix commutes with $\varphi(x)$ for all $x\in L$. This can be done by showing $a_{ij}=-b_{ji}$

	Then in computing $[xc_\varphi]$, we find that the coefficients of $\varphi(x_i)\varphi(y_j)$
	is $a_{ij}+b_{ji}=0$. Thus $c_\varphi$ commutes with the matrices of $\varphi(L)$.

	Hence if $V=k^n$ ws te module being acted upon and if $V$ is irreducible, then $c_\varphi$ must be a scalar (Schur).
	What is this scalar? Well since $\tr c_\varphi=m=\dim L$, we get
	\[\frac{\dim L}{\dim V}I=\frac{m}{n}I.\]
\end{rmk}
\begin{prf}
	The proof here will basically follow Humphreys, but will include a comment that \textit{should have appeared} in the text
	(much to Monty's chagrin).

	Notice that we must show that every submodule $N$ of $M$ has a complement $N'$ so that $M$ is the direct sum of $N$ and $N'$ and $N'$ is a submodule.

	For the rest of it, peek at the book. :)
\end{prf}
\begin{rmk}
	This result fails for semisimple Lie algebras in positive characteristic, although we won't really be dealing with this much.
	This parallels the fact that Maschke's theorem fails when $\ch k$ divides the order of $G$.
\end{rmk}
\subsection{Structure}
We now consider the simples case of a semisimple Lie algebra, namely $\sl(2,k)$. Here we know that we have the basis
$h=e_{11}-e_{22}$, $x=e_{12}$ and $y=e_{21}$ where
\[[hx]=2x\quad [hy]=-2y\quad [xy]=h\]

Next time we will see that the diagonal matrix $hin L$ continues to act diagonally on \textit{any} finite dimensional representation of $L$.
Thus the first thing to look at here is how the representation $V$ decomposes as a sum of its eigenspaces
\[V=\oplus_{\lambda\in k}V_\lambda\]
but since $V$ is assumed to be finite dimensional, this is actually a finite sum. In particular, there is $\lambda$ such that 
$V_{\lambda+2}=0$.
\begin{defn}
	We call the $\lambda$ where $V_\lambda\ne 0$ \textbf{weights} of $V$.
\end{defn}
So there is a weight $\lambda$ such that $\lambda+2$ is not a weight. In general, due to the relations above on $L$,
\[xV_\mu\subseteq V_{\mu+2}\quad yV_\mu\subseteq V_{\mu-2}\]
and we call $x$ and $y$ (or their matrices) the \textbf{raising and lowering operators}.

Then we can pick $\lambda$ a weight such that $\lambda+2$ is not and pick $v_0\in V\setminus 0$. Then set
\[v_i=\frac{y^iv_0}{i!}\]
so 
\[hv_i=(\lambda-2i)v_i,\quad yv_i=(i+1)v_{i+1}\quad xv_i=(x-i+1)v_{i-1}\]
where this can be proved by induction.

But the nonzero vectors $v_i$ are independent, so there is a least $m$ with $v_m\ne 0$ and $v_{m+1}=0$. 
The last equation with $i=m+1$ then forces $\lambda-m=0$, so $\lambda=m$ is a nonnegative integer!

It follows that if $V$ is irreducible that $\dim V=m+1$ and $v_0,\dots,v_m$ are a basis for $V$.

This ends up capturing the situation with \textit{any} irreducible representation of $\sl(2,k)$. We get integer eigenvalues, all of the same parity, from $-m$ to $m$.

\section{January 28, 2019}
We'll start off with tying up some loose ends from last time: 
\begin{lem}
	Let $L$ be a semisimple subalgebra of $\gl(n,k)$ (once more, $k=\bar k$ and $\ch k=0$). Let $x\in L$. Then 
	the semisimple and nilpotent parts $x_s$ and $x_n$ of $x$ lie in $L$. Thus the Jordan decompositions of $x$ 
	(as a matrix and as an element in $L$) coincide.
\end{lem}
\begin{prf}
	To see this, recall that we know that $\ad x_x$ and $]ad x_n$ are polynomials with zero constant term in $\ad x$ and
	so bracket $L$ into $\gl(V)$. So if $x_s,x_n\in N=\{y\in\gl(n,k):[yL]\subseteq L\}$ and $L$ is an ideal of $N$.

	We want to show that $L-N$ but this is false! we need to get the scalar matrices out. So replace $N$ by those elements 
	such that if $y\in N$ then $y$ restricted to any irreducible submodule $k^n$ of $L$ has trace zero.

	We have $L\subseteq N$... (snip)

	So by Weyl's thoerem we can write $N=L\oplus M$ as an $L$ module. But $L$ is an ideal of $N$ and $[LM]=0$, so the matrices in $M$
	all commute with $L$. By Schur's lemma, they act by scalars on any irreducible submodule $W$ of $k^n$, which must be zero, by the trace condition.

	Since $k^n$ is the direct sum of irreducible modules, we get $M=$, so $L=N$.
\end{prf}

Hence we have fully justified our characterization of irreducible $(m+1)$-dimensional $\sl(2,k)$ modules $V$:
\begin{itemize}
	\item We always have a basis $v_0,\dots,v_m$ with $v_{-1}=v_{m+1}=0$ such that:
	\item $hv_i=(m-2i)v_i$
	\item $yv_i=(i+1)v_{i+1}$
	\item $xv_i=(m-i+1)v_{i-1}$
\end{itemize}
we can check that the bcracket relations are satisfied starting from this as a definition.
\begin{rmk}
	In fact, if you replace $m$ by any $\lambda\in k$ and let all $v_i$'s form a basis for all non-negative $i$ (that is easing
	the restriction $v_{m+1}=0$), you get an infinite dimensional $L$-module. Whenever $\lambda\in\N$, this module is \textbf{not}
	completely reducible.
\end{rmk}
\subsection{One more $\sl(2,k)$ Computation}
\[(\exp\ad ax)(\exp\ad hy)(\exp\ad ax)=\begin{pmatrix}
	1+ab & a(2+ab)\\
	b & 1+ab
\end{pmatrix}=\begin{pmatrix}
	0 & -b^{-1}\\
	b & 0
\end{pmatrix}\]
if $ab=-1$. Multiplying by its transpose, we get
\[\begin{pmatrix}
	b^{-2} & 0 \\
	0 & b^2
\end{pmatrix}\]
So whenever $K$ is algebraically closed, we get a set of generators of $\sl(2,k)$ as products of matrices in $\Int L$ which is therefore 
isomorphic to $\operatorname{PSL}(2,k)$.

The action of $L$ on any irreducible module $V$ always lifts to $\operatorname{SL}(2,k)$ and lifts to $\operatorname{PSL}(2,k)$ if and only if $\dim V$ is odd.

The moral to draw from our classification of $\sl(2,k)$ moduoes of finite dimension is that the weights ($h$-eigenvalues) always occur in strings
of integers. In particular the weight $i$ occurs overall with the same multiplicity (eigenspace dimension) as $-i$.

Also for an irreducible module either the weight zero or the weight 1 occurs with multiplicity one (but not both).

\subsection{Return to semisimple Lie algebras}
Consider decomposing such an $L$ as a module over itself via the adjoint representation. If $L$ had only nilpotent elements, it itself would be nilpotent (it isn't)
so $L$ has a \textbf{toral subalgebra} $T$ consisting of only semisimple elements.

Then, in fact, $T$ is abelian, since otherwise we would have $[xy]=ky$ for $x,y\in T$ and $k\ne 0$.
But on the other hand, since $\ad y$ is semisimple, but on the other hand $[yx]$ is a sum of eigenvectors of $\ad y$
each with nonzero eigenvalues, a contradiction.

\subsection{Decomposing $L$}
Now let's fix a \textbf{maximal toral subalgebra} $\h$ (that is, maximal among toral subalgebras). Then we have
\[L=\oplus_{\alpha\in\h^*}L_\alpha\]
where $L_\alpha=\{x\in L:[xh]=\alpha(h)x\}$, since any commuting set of semisimple (diagonalizable) matrices are simultaneously diagonalizable.

Now by Jacobi, $[L_\alpha L_\beta]\subseteq L_{\alpha+\beta}$ and bringing back in the Killing form,
\[\kappa(L_\alpha,L_\beta)=0,\quad\text{if}\quad \alpha+\beta\ne 0\]
so that $\kappa$ is nondegenerate on $L_0$:
\[\kappa(L_0,L_\alpha)=0\quad\forall \alpha\ne 0.\]

\begin{lem}
	$H=L_0$, or equivalently, $H$ is its own centralizer.
\end{lem}

The proof can be found in the book, but using this, we can rewrite our decomposition as
\[L=\h\oplus\bigoplus_{\alpha\in\h^*}L_\alpha\]
and this is called the \textbf{root space decomposition} of $L$ in terms of $\alpha\in\h^*$.
The terms $\alpha\in\h^*$  occurring above are called the \textbf{roots of $L$} (relative to $\h$). By definition \textbf{no root is zero.}

Notation: Write $\Phi$ to be the set of all (nonzero) roots in $\h^*$.
